---
title: "Functional Enrichment Analysis of Candidate Genes (Corrected)"
author: "Daniel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

### 1. Setup and Data Loading

```{r libraries}
# En un chunk setup de tu .Rmd:
if (!requireNamespace("BiocManager", quietly=TRUE)) install.packages("BiocManager")
pkgs <- c("clusterProfiler","ggplot2","dplyr","GO.db","knitr")
for (p in pkgs) {
  if (!requireNamespace(p, quietly=TRUE)) BiocManager::install(p)
}
# Cargar en orden:
library(GO.db)         # primero GO.db (AnnotationDbi)
library(clusterProfiler)
library(ggplot2)
library(dplyr)         # luego dplyr, sobreescribe select()
library(knitr)

```

---

### 2. Loading and Preparing Data

We now load our small list of candidate genes and the **new, complete background universe** file.

```{r load-data}
# Lista de genes candidatos
candidate_genes <- scan("final_gene_list.txt", what=character(), sep="\t")
message("Genes candidatos: ", length(candidate_genes))

# Data frame background completo
full_bg_df <- read.delim("gene2go_universe_full.tsv", header=FALSE, col.names=c("gene","go_id"))
message("Genes en universo: ", length(unique(full_bg_df$gene)))

# Extraer columnas con dplyr::select SIN conflicto
term2gene <- full_bg_df %>% dplyr::select(go_id, gene)

# Preparar TERM2NAME: mapea GO IDs a nombres legibles
go_ids <- unique(term2gene$go_id)
term2name <- data.frame(
  go_id = go_ids,
  name  = Term(go_ids)      # Term() de GO.db
) %>% filter(!is.na(name))
```

---

### 3. Performing GO Enrichment Analysis

With the correct universe, we can now perform the statistical test for all three ontologies.

```{r enrichment}
go_results <- list()
for (ont in c("BP","CC","MF")) {
  message("Ontología: ", ont)
  # Filtrar términos de esta ontología
  ont_terms <- term2name %>% filter(Ontology(go_id)==ont)
  ont_t2g   <- term2gene   %>% filter(go_id %in% ont_terms$go_id)

  res <- enricher(
    gene      = candidate_genes,
    universe  = unique(full_bg_df$gene),
    TERM2GENE = ont_t2g,
    TERM2NAME = ont_terms
  )
  go_results[[ont]] <- res
  message("→ ", if(is.null(res) || nrow(res@result)==0) "0" else nrow(res@result),
          " términos significativos")
}
```

---

### 4. Visualizing Results

Finally, we display the results for each ontology that produced significant hits.

```{r results-visualization, fig.width=12, fig.height=10}
for (ont in names(go_results)) {
  res <- go_results[[ont]]
  if (!is.null(res) && nrow(as.data.frame(res))>0) {
    cat("\n### Ontología:", ont, "\n")
    print(kable(as.data.frame(res), caption=paste("Enriquecidos", ont)))
    print(dotplot(res, showCategory=20) + ggtitle(paste("Top",ont)))
  }
}
```