---
title: "Functional Enrichment Analysis of Candidate Genes in 'Búlida Precoz'"
author: "Daniel González Palazón"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

### 1. Introduction and Environment Setup

This document performs a complete functional analysis of the final candidate genes. The objective is to characterize these genes and determine if they are involved in relevant biological processes through a Gene Ontology (GO) enrichment analysis.

**This is the first code block and it is crucial. It ensures that all necessary libraries, including `dplyr`, are installed and loaded before the rest of the analysis is executed.**


```{r libraries, message=FALSE, warning=FALSE}
# Install BiocManager if not available
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")

# Install and load clusterProfiler for enrichment analysis
if (!requireNamespace("clusterProfiler", quietly = TRUE)) BiocManager::install("clusterProfiler")
library(clusterProfiler)

# Install and load GO.db for GO annotations
if (!requireNamespace("GO.db", quietly = TRUE)) BiocManager::install("GO.db")
library(GO.db)

# Install and load dplyr for data manipulation
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
library(dplyr)

# Install and load ggplot2 for plotting
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
library(ggplot2)

# Install and load tidyr for data tidying
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
library(tidyr)
```

---

### 2. Data Loading and Preparation

Now we will load the annotation data we have prepared and format it for enrichment analysis.

```{r load-data}
# Define file paths
blast_file <- "blast_with_GO.txt"
desc_file  <- "accession_descriptions.txt"
bg_file    <- "gene2go_background.txt"

# Read BLAST results with GO assignments
blast_df <- read.delim(blast_file,
                       header = TRUE,
                       sep = "\t",
                       stringsAsFactors = FALSE)

# Read accession descriptions
desc_df <- read.delim(desc_file,
                      header = TRUE,
                      sep = "\t",
                      stringsAsFactors = FALSE)

# Read background gene-to-GO mappings
bg_df <- read.delim(bg_file,
                    header = FALSE,
                    sep = "\t",
                    stringsAsFactors = FALSE)
colnames(bg_df) <- c("gene", "GO_ID")
```

---

### 3. Characterization of Candidate Genes

We present a summary table with the functional annotation for each gene of interest, based on its best BLASTp hit.

```{r summary-table}
# Debug: check blast_df
print(paste("BLAST rows:", nrow(blast_df)))
print("Columns in blast_df:"); print(names(blast_df))

# Split GO_IDs column into a list
blast_df2 <- blast_df %>%
  mutate(GO_IDs_list = strsplit(GO_IDs, ";\\s*"))

# Debug: view first entries of GO_IDs_list
print(head(blast_df2$GO_IDs_list))

# Unnest GO terms and rename to GO_ID
blast_unnest <- blast_df2 %>%
  tidyr::unnest(GO_IDs_list) %>%
  rename(GO_ID = GO_IDs_list)

# Debug: check number of rows after unnest
print(paste("Unnested rows:", nrow(blast_unnest)))

# Merge with descriptions and summarize by qseqid
summary_df <- blast_unnest %>%
  left_join(desc_df, by = "accession") %>%
  group_by(qseqid) %>%
  summarise(
    Best_hit    = first(accession),
    Description = first(Description),
    pident      = first(pident),
    evalue      = first(evalue),
    GO_IDs      = paste(unique(GO_ID), collapse = "; ")
  ) %>%
  ungroup()

# Debug: inspect summary_df
print(paste("Summary rows:", nrow(summary_df)))
print(head(summary_df))

# Display summary table
knitr::kable(summary_df, caption = "Summary of Candidate Genes (UniProt IDs)")

```

---

### 4. GO Enrichment Statistical Analysis

We will perform the overrepresentation test for the three main GO ontologies: Biological Process (BP), Molecular Function (MF), and Cellular Component (CC).

```{r enrichment-analysis}
# Debug: ensure blast_unnest exists and has data
message("Rows in blast_unnest: ", nrow(blast_unnest))

# Create frequency table of GO terms
library(dplyr)
go_freq <- blast_unnest %>%
  filter(!is.na(GO_ID)) %>%
  count(GO_ID, name = "Count") %>%
  arrange(desc(Count))

# Join GO term descriptions from GO.db
go_desc <- AnnotationDbi::select(
  GO.db,
  keys    = go_freq$GO_ID,
  columns = c("GOID","TERM"),
  keytype = "GOID"
) %>%
  rename(GO_ID = GOID, Term = TERM)

go_freq <- left_join(go_freq, go_desc, by = "GO_ID")

# Debug: show top 10 frequencies
print(head(go_freq, 10))

# Plot barplot of top 10 GO terms
top10 <- go_freq %>% slice_max(Count, n = 10)

library(ggplot2)
ggplot(top10, aes(x = reorder(Term, Count), y = Count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = "Top 10 GO Terms by Frequency",
    x     = "GO Term",
    y     = "Number of Genes"
  )
```
